---
title: "DiscreteCharacters"
author: "Brian O'Meara"
date: "2/16/2017"
output: html_document
---

For this exercise, explore looking at discrete character models.

Remember for homework, change `eval=TRUE` for all the R blocks, debug, commit, and do a pull request.

These are useful packages in this area (far from exhaustive list).

```{r, eval=TRUE}
library(ape)
library(geiger)
library(phytools)
library(phangorn)
library(corHMM)
```

You'll need to get data into R in some way. 

```{r, eval=TRUE}
tree <- read.tree("Mac_.tre")
discrete.data <- read.csv(file="Mac_raw2.csv", stringsAsFactors=FALSE) #death to factors.
```

Data are often not right in some way. They might not match the taxa in your tree, there may be missing data, etc. `geiger::treedata` is a great function for getting a tree and data that match, but your data may need other cleaning. Do it as a function so it's repeatable.

```{r, eval=TRUE}
CleanData <- function(tree, discrete.data) {
  tree$tip.label <- gsub("_", " ", tree$tip.label)
  rownames(discrete.data) <- discrete.data[,1]
  # make it a character matrix
  discrete.data <- as.matrix(discrete.data[,-1])
  # define ambiguous states "N" and "?"
  data <- phyDat(as.matrix(discrete.data), type="USER", levels=c("0", "1", "2"), ambiguity = c("N", "?"))
  # reorder data
  data <- subset(data, tree$tip.label) 
  list(phy=tree, data = data)
}

cleaned.results <- CleanData(tree, discrete.data)
cleaned.tree <- cleaned.results$phy
cleaned.data <- cleaned.results$data

```

It's critically important to LOOK at what you have. Are there weird values? Has the match between taxa and state gone correctly? Do you think you have binary data, but there's actually only state 1? Especially as data sets grow (yay), and are assembled using scripts rather than by error-prone, non-reproducable hands (double yay), scientists are increasingly less likely to deeply look at our data. That's bad -- don't be that person.

```{r, eval=TRUE}
VisualizeData <- function(tree, data) {
  plot(tree)
  print(data)
}

discrete.data.matrix <- as.matrix(cleaned.data)
rownames(discrete.data.matrix) <- rownames(cleaned.data)
print(discrete.data.matrix)
```

First, let's use parsimony to look at ancestral states:

```{r, eval=TRUE}
anc.p <- ancestral.pars(cleaned.tree, cleaned.data)
print(anc.p)

plotAnc(tree, anc.p, attr(anc.p, "index")[1])
  ###Error in data[[i]][rows, ] : subscript out of bounds

plotAnc(tree, anc.p, attr(anc.p, "index")[15])
  ###Error in data[[i]][rows, ] : subscript out of bounds
```

Do you see uncertainty? What does it mean?

Now, plot the likelihood estimates. 

```{r, eval=TRUE}
anc.ml <- ancestral.pml(pml(tree, cleaned.data), type="ml")
#Error in pml(tree, cleaned.data) : tree must have edge weights
plotAnc(tree, anc.ml, 1)
```

How does this differ from parsimony? 

Why does it differ from parsimony?

What does uncertainty mean?

Now, to the biological questions. For many of these, `corHMM` will be a useful package.

1. How can you estimate transition rates between states? Do it.
2. How could you examine if transition rates are equal?
3. Think about the Lewis (2001) MKV model. Are your traits all variable? Will using this make sense for your data? Try using it. Do results change?
4. How could you test order of state evolution?